Code used to generate paper results : Database comparison and CT characteristics

### 2023-11-16 Paper_code-Database_CT.Rmd
worked on coactivation in tumors

```{r}
new <- CT_genes %>% 
  filter(!external_gene_name%in%common) %>% 
  pull(external_gene_name)

length(new)

# Charles' idea
# TCGA_expression(tumor = "all", genes = new, units = "log_TPM")
```

@@

@@

>>>>

```{r, warning = FALSE}
new <- CT_genes %>% 
  filter(!external_gene_name%in%common)

new

table(new$testis_specificity)
table(new$X_linked, new$regulated_by_methylation)
TCGA_expression(tumor = "all", genes = new$external_gene_name, 
                units = "log_TPM")
TCGA_expression(tumor = "all", 
                genes = filter(new, X_linked & regulated_by_methylation)$external_gene_name, 
                units = "log_TPM")
```
There are `r dim(new)[1]` new CT genes in CTexploreR. These are mainly testis
specific and on autosomes. Regulation by methylation is not the majority of 
them. This makes sens as other databases have been focused on regulation by 
methylation, so as here we don't segregate with that we can find these. There 
is only 11 new "major" CT that are on the X chromosome and regulated by 
methylation. CT45 are not that new.

Expression in tumours doesn't strike that much.

@@

Next graph was removed from paper

@@

### Coactivation in tumours

We know that CT genes that are regulated by methylation are activated in 
tumours due to methylation. However, those that are not it is more blurry. It 
is interesting to look at their expression to see if there are groups of genes 
that are activated together in tumours which would suggest another activation :
a transcription factor ? a gene expresion program from meiosis ? timing of 
expression ? function ?

```{r not_reg}
not_regulated <- filter(CT_genes, !regulated_by_methylation)$external_gene_name
```

As they can blur the clustering, I sometimes remove genes that are in the 
background and  those that are expressed a lot.

- BRCA :

```{r BRCA}
TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
                tumor = "BRCA", units = "log_TPM")
# Removing genes to see clearer

BRCA_genes <- rowData(TCGA_TPM) %>% 
  as_tibble() %>% 
  filter(external_gene_name %in% not_regulated) %>% 
  filter(percent_pos_BRCA > 0.01 & percent_pos_BRCA < 30) %>% 
  pull(external_gene_name)

BRCA_genes_ordered <- 
  BRCA_genes[row_order(draw(TCGA_expression(BRCA_genes, tumor = "BRCA", 
                                            units = "log_TPM")))]

BRCA_cluster1 <- BRCA_genes_ordered[c(which(BRCA_genes_ordered == "AKAP14"):
                       which(BRCA_genes_ordered == "C11orf97"))]
BRCA_cluster2 <- BRCA_genes_ordered[c(which(BRCA_genes_ordered == "GTF2I-AS1"):
                       which(BRCA_genes_ordered == "SLIT3-AS1"))]

We can see two small clusters.

- LUAD :

```{r LUAD}
TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
                tumor = "LUAD", units = "log_TPM")
# Removing genes to see clearer
LUAD_genes <- rowData(TCGA_TPM) %>% 
  as_tibble() %>% 
  filter(external_gene_name %in% not_regulated) %>% 
  filter(percent_pos_LUAD > 0.01 & percent_pos_LUAD < 30) %>% 
  pull(external_gene_name)

LUAD_genes_ordered <- 
  LUAD_genes[row_order(draw(TCGA_expression(LUAD_genes, tumor = "LUAD", 
                                            units = "log_TPM")))]

LUAD_cluster1 <- c("TSPY1", "TSPY2")
LUAD_cluster2 <- c("SPANXC", "SPANXB1",  "SPANXD")
LUAD_cluster3 <- LUAD_genes_ordered[c(1:22)]
LUAD_cluster4 <- LUAD_genes_ordered[c(which(LUAD_genes_ordered == "GTF2I-AS1"):
                       which(LUAD_genes_ordered == "SLIT3-AS1"))]
```

22 first genes are clearly quite clustering, there is also a small SPANX cluster,
a small TSPY cluster ans a bigger one around the middle of the graph.

- LUSC :

```{r LUSC}
TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
                tumor = "LUSC", units = "log_TPM")
# Removing genes to see clearer

LUSC_genes <- rowData(TCGA_TPM) %>% 
  as_tibble() %>% 
  filter(external_gene_name %in% not_regulated) %>% 
  filter(percent_pos_LUSC > 0.01 & percent_pos_LUSC < 30) %>% 
  pull(external_gene_name)

LUSC_genes_ordered <- 
  LUSC_genes[row_order(draw(TCGA_expression(LUSC_genes, tumor = "LUSC", 
                                            units = "log_TPM")))]

LUSC_cluster1 <- c("TSPY1", "TSPY2",  "TSPY9", "TSPY10")
LUSC_cluster2 <- c("SPANXC", "SPANXB1",  "SPANXD")
LUSC_cluster3 <- LUSC_genes_ordered[c(1:15)]

We can see a cluster at the top and two smaller ones corresponding to SPANX and
TSPY genes. As well as quite a big cluster with the 15 first lines

- SKCM :

```{r SKCM}
SKCM_genes <- rowData(TCGA_TPM) %>% 
  as_tibble() %>% 
  filter(external_gene_name %in% not_regulated) %>% 
  filter(percent_pos_SKCM > 0.01 & percent_pos_SKCM < 30) %>% 
  pull(external_gene_name)

TCGA_expression(filter(CT_genes, !regulated_by_methylation)$external_gene_name,
                tumor = "SKCM", units = "log_TPM")
# Lots of genes activated almost everywhere or nowhere that can mask clusters
TCGA_expression(SKCM_genes,
                tumor = "SKCM", units = "log_TPM")

SKCM_cluster <- c("TSPY1", "TSPY2", "TSPY3", "TSPY9", "TSPY10")
```

It is clear that there is a TSPY cluster again.

- Comparison

TSPY and SPANX genes are found reactivated/coactivated in different types of 
tumours. Let's compare the bigger groups found in LUAD, LUSC and BRCA

```{r upste_clusters}
upset(fromList(list(LUSC = LUSC_cluster3,
               LUAD1 = LUAD_cluster3,
               LUAD2 = LUAD_cluster4,
               BRCA1 = BRCA_cluster1,
               BRCA2 = BRCA_cluster2)))
```

There is quite a lot of similarities here between the clusters.

Let's have a look at the timings of expression.

```{r timing_cluster, message = FALSE}
testis_expression(c("TSPY1", "TSPY2", "TSPY3", "TSPY9", "TSPY10"), 
                  cells = "all")
testis_expression(c("SPANXC", "SPANXB1",  "SPANXD"), 
                  cells = "all")
testis_expression(LUSC_cluster3, cells = "all")
testis_expression(LUAD_cluster3, cells = "all")
testis_expression(LUAD_cluster4, cells = "all")
testis_expression(BRCA_cluster1, cells = "all")
testis_expression(BRCA_cluster2, cells = "all")
```

They mainly seem to, in a cluster, be expressed at the same time.
